---
title: "Manual vs. Automatic Transmissions Related to MPG"
author: "Neal S"
output: pdf_document
---
```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(dev = 'pdf')
#opts_chunk$set(dev="png", 
#               dev.args=list(type="cairo"),
#               dpi=96)
#opts_chunk$set(dev="tiff", 
#               dev.args=list(compression="lzw"),
#               dpi=300)
```
# Abstract
In this paper we will explore the relationship between automobile transmission
type (manual vs. automatic) and miles per gallon (MPG). We will use data from
the _mtcars_ dataset provided with R. The accepted rule in the industry is
that manual transmissions allow for better MPG ratios. We prove that notion
to be true using regression analysis on the provided dataset.

# The Dataset
The dataset we use is provided with the standard R installation and is taken
from _Motor Trend_ magazine in 1974. It contains data from 32 automobiles
contemporary to the issue and details 10 aspects of their design and performance
along with a fuel consumption metric.

Below are two plots that show some aspects of the dataset as they relate to MPG.
The first shows the MPG reading for each car in the dataset. The next plot shows
MPG as it relates to overall car weight. It also highlights which points refer
to automatic vs. manual transmissions.

```{r, echo=FALSE, cache=TRUE, fig.width=3.5, fig.height=3}
library(ggplot2)
data(mtcars)

ggplot(mtcars, aes(x=seq_along(mpg),y=mpg)) +
    geom_point(aes(color=as.factor(cyl), size=3)) +
    ylab('Miles per Gallon') + xlab('Data Item') +
    scale_color_discrete(name='# of Cylinders') +
    scale_size(guide=FALSE) +
    ggtitle('Miles Per Gallon for Each Car')

mtc<-mtcars
mtc$tran <- 'Automatic'
mtc$tran[mtc$am==1] <- 'Manual'
ggplot(mtc, aes(x=wt, y=mpg)) +
    geom_point(aes(color=tran, size=3)) +
    geom_smooth(method='lm') +
    xlab('Car Weight (in 1000 lbs)') +
    ylab('Miles per Gallon') +
    scale_size(guide=FALSE) +
    scale_color_discrete(name='Transmission') +
    ggtitle('Miles per Gallon vs. Car Weight')



```

From the figures you can see that MPG is related to more than simply the
transmission type of a car. The weight of the car and number of cylinders
in the engine appear to have some influence on MPG as well. We will explore
all of these issues further.


# Analysis

Briefly mention other metrics related to MPG such as cylinders, weight, and
displacement. also make a note of how accurate the data actually is or is not,
considering that we only have 32 entries.


## Models

Need to fit multiple models and show which one we like and why (WTF?)

Should do a residual plot and explain some diagnostics on the data as to
why it doesn't fit exactly.

## Transmission Type vs. MPG

Show the results. Is this valid? We should try to compare apples-to-apples
and show that, e.g., holding other vars fixed, transmission type does make
a difference (for 4-cyl cars, manual is better, etc. or for displacement
groups, or rear axle ratio groups, etc.)

# Conclusion

Need to put a qualifier in there that shows our uncertainty and why.

