---
title: "Manual vs. Automatic Transmissions Related to MPG"
author: "Neal S"
header-includes: \usepackage{graphicx}
graphics: yes
output:
    pdf_document:
        fig_caption: yes
---
```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(grid)
data(mtcars) # am:1=manual, vs:1=straight(inline)
opts_chunk$set(dev = 'pdf')
#opts_chunk$set(dev="png", 
#               dev.args=list(type="cairo"),
#               dpi=96)
#opts_chunk$set(dev="tiff", 
#               dev.args=list(compression="lzw"),
#               dpi=300)
source('multiplot.R')
```
# Abstract
In this paper we will explore the relationship between automobile transmission
type (manual vs. automatic) and miles per gallon (MPG). The generally accepted
rule in the automobile industry is
that manual transmissions allow for better MPG ratios. We prove that notion
to be true using regression analysis on our dataset. We will use
data from
the _mtcars_ dataset provided with R.

# The Dataset
The dataset we use is provided with the standard R installation and is taken
from _Motor Trend_ magazine in 1974. It contains data from 32 automobiles
contemporary to the issue and details 10 aspects of their design and performance
along with a fuel consumption metric.

Figure 1 contains two plots that show some aspects of the dataset as they
relate to MPG.
The first shows the MPG value for each car in the dataset, highlighting
the number of cylinders in the engine of that car. The next plot shows
MPG as it relates to overall car weight. The color highlights which points
refer to automatic vs. manual transmissions.

```{r fig1, echo=FALSE, fig.width=12, fig.height=5, fig.cap="A quick look at the dataset. You can see from these two plots that MPG is related to other variables besides just transmission type.", out.height="2in", out.width="5in", fig.align='center'}
p1<-ggplot(mtcars, aes(x=seq_along(mpg),y=mpg)) +
    geom_point(aes(color=as.factor(cyl), size=3)) +
    ylab('Miles per Gallon') + xlab('Data Item') +
    scale_color_discrete(name='# of Cylinders') +
    scale_size(guide=FALSE) +
    ggtitle('Miles Per Gallon for Each Car')

mtc<-mtcars
mtc$tran <- 'Automatic'
mtc$tran[mtc$am==1] <- 'Manual'
p2<-ggplot(mtc, aes(x=wt, y=mpg)) +
    geom_point(aes(color=tran, size=3)) +
    geom_smooth(method='lm') +
    xlab('Car Weight (in 1000 lbs)') +
    ylab('Miles per Gallon') +
    scale_size(guide=FALSE) +
   scale_color_discrete(name='Transmission') +
    ggtitle('Miles per Gallon vs. Car Weight')

multiplot(p1, p2, cols=2)
```

From these plots you can see that MPG is influenced by more than simply the
transmission type of a car. The weight of the car and number of cylinders
in the engine appear to have some influence on MPG as well. We will now explore
all of these issues further.

# Analysis of the dataset

To fully examine the relationship of transmission type to MPG, we must also
look at the other variables as well. This is simply because
relationships might exist between transmission type and some of the other
variables. If a relationship exists, it would confound the results obtained
by ignoring the other variables. That is, any results would be less meaningful.

So we will build models that include other variables to test how they
change what we see as the relationship between transmission type and MPG.

The dataset includes a few variables that are measurements rather than
physical characteristics of the automobile design, MPG being one of them.
So our model need not include any variables as predictors if they are
measurements (outcome). These variables are _MPG_, _qsec_, and _hp_.
We will ignore the last two and of course use **MPG** as our outcome.

## The Models

First, the easy and simple case. Using only the transmission type as a
predictor variable, we construct a simple linear model:
```{r models, cache=TRUE}
m1 <- lm(mpg ~ am, data=mtcars)
```
The resulting model fit is simply the mean MPG values
for the two groups: those cars with and without automatic transmission.
Manual transmission average MPG is `r round(m1$coef[1] + m1$coef[2],2)` and
automatic transmission average MPG is `r round(m1$coef[1],2)`.
This would appear to
answer our original question concerning MPG and transmission type,
but we can't really stop here; that would be irresponsible data science.
There's far more to the data than those two variables.

```{r models2, cache=TRUE}
m2 <- lm(mpg ~ am + wt + cyl, data=mtcars)
m3 <- lm(mpg ~ am + disp + wt + vs + carb, data=mtcars)
m4 <- lm(mpg ~ am + drat + wt +  carb, data=mtcars)
m5 <- lm(mpg ~ am + wt + cyl + vs, data=mtcars)

summary(m2)$coeff
summary(m3)$coeff
summary(m4)$coeff
summary(m5)$coeff
```
From looking at the resulting model coefficients, we can easily see
that transmission
type, _am_, always has a positive relationship to MPG. That is, its
resulting model coefficient is greater than zero for all
five models. So a higher value of _am_ means higher (better) MPG.
This translates to a manual transmission (where _am=1_)
offering better MPG than an automatic transmission (where _am=0_).

...Need to look at how much auto vs manual trans really changes things...

As a final sanity check, in Figure 2
we look at the residual plot and see that our data is indeed a good
candidate for using linear regression. The random pattern of residuals
spread evenly with respect to the horizontal line 0 indicates so.

```{r resid, echo=FALSE, fig.width=8, fig.height=6, out.width='4in', fig.align='center', fig.cap='The residuals from model 4 appear to be evenly spread around 0, indicating that this dataset is a good candidate for a linear model.'}
plot(resid(m4))
abline(h=0, col='red', lty=2)
```


# Conclusion and Assumptions

Need to put a qualifier in there that shows our uncertainty and why.

So in answer to the question
“Is an automatic or manual transmission better for MPG” we agree with
the automobile industry's notion that manual transmissions provide
better MPG.

However, it should be noted that while transmission type certainly
affects overall MPG, it is certainly
not the only predictor. Other variables related to automobile design
can have a very large impact on MPG as well.

"Quantify the MPG difference between automatic and manual transmissions"
Figure 3 shows a simple box plot of MPG for automatic vs manual
transmissions in our dataset. From this data, one can easily see that
a manual transmission can, on average, make more than a 40% difference
in MPG for the better.

Another caveat is that there are only 32 samples. This is most likely not
a large enough population to formulate a general rule regarding automobiles
contemporary to 1974. It does, however, make a nice academic exercise and
that's what we're doing here.

```{r box, echo=FALSE, fig.width=6, fig.height=6, out.width='3in', fig.align='center', fig.cap='The average increase of MPG for a manual transmission vs. an automatic one is over 40\\%.'}
mtc <- mtcars[,c('am', 'mpg')]
mtc$am <- as.factor(mtc$am)
levels(mtc$am)[levels(mtc$am)==0] <- 'Automatic'
levels(mtc$am)[levels(mtc$am)==1] <- 'Manual'
names(mtc)[names(mtc)=='am'] <- 'Transmission'
ggplot(mtc, aes(factor(Transmission), mpg, fill=Transmission)) +
    geom_boxplot() +
    ylab('Miles per Gallon') +
    xlab('Transmission Type') +
    theme(legend.position="none") +
    ggtitle('MPG vs. Transmission Type')
```

